Spatial variation of the native colon microbiota in healthy adults

### 2018-02-12 submission/manuscript.Rmd
reviewer edits completed

---
title: "Spatial variation of the native colon microbiota in healthy adults"
csl: aacr.csl
fontsize: 12pt
output:
  pdf_document:
    includes: null
    keep_tex: yes
  html_document: default
  word_document: default
sansfont: Arial
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
- \renewcommand{\familydefault}{\sfdefault}
- \usepackage{graphicx}
geometry: margin=1.0in
bibliography: references.bib
---
\vspace{8mm}

@@

Word count: 3,953/5,000

@@

```{r datasets, echo=FALSE}
meta <- '../data/raw/kws_metadata.tsv'
shared <- '../data/mothur/kws_final.an.shared'
tax <- '../data/mothur/kws_final.an.cons.taxonomy'
subsample <- read.table(file='../data/mothur/kws_final.an.0.03.subsample.shared', header=T)

meta_file <- read.table(file='../data/raw/kws_metadata.tsv', header = T)
shared_file <- read.table(file='../data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
tax_file <- read.table(file='../data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')

@@

```{r randomforest, echo=FALSE}

#update this to just be importing end files off of optimization scripts to pull AUCs from

@@

The microbiome has been implicated in the development of colorectal cancer (CRC) and inflammatory bowel diseases (IBD). The specific traits of these diseases vary along the axis of the digestive tract. Further, variation in the structure of the gut microbiota has been associated with both diseases. Here we profiled the microbiota of the healthy proximal and distal mucosa and lumen to better understand how bacterial populations vary along the colon. We used a two-colonoscope approach to sample proximal and distal mucosal and luminal contents from the colons of 20 healthy subjects that had not undergone any bowel preparation procedure. The biopsies and home-collected stool were subjected to 16S rRNA gene sequencing and Random Forest classification models were built using taxa abundance and location to identify microbiota specific to each site. The right mucosa and lumen had the most similar community structures of the five sites we considered from each subject. The distal mucosa had higher relative abundance of _Finegoldia, Murdochiella, Peptoniphilus, Porphyromonas_ and _Anaerococcus_. The proximal mucosa had more of the genera _Enterobacteriaceae, Bacteroides_ and _Pseudomonas_. The classification model performed well when classifying mucosal samples into proximal or distal sides (AUC=0.850). Separating proximal and distal luminal samples proved more challenging (AUC=0.580) and specific microbiota that differentiated the two were hard to identify. By sampling the unprepped colon, we identified distinct bacterial populations native to the proximal and distal sides. Further investigation of these bacteria may elucidate if and how these groups contribute to different disease processes on their respective sides of the colon.

Words: 259/260

@@

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that were distinct at each site, we constructed several Random Forest models trained using OTU relative abundances. We built the first model to classify the luminal versus mucosal samples for the proximal and distal sides, independently (Fig. 4A). The models performed well when classifying these samples (proximal AUC = 0.764, distal AUC = 0.908). The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the proximal lumen and mucosa, OTUs affiliated with the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and _Enterobacteraceae_ were included in the best model (Fig. 5A). The model to differentiate between the distal lumen and mucosa included OTUs affiliated with the _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and _Anaerococcus_ (Fig. 5B). These results indicated that there were fine differences between the different sites of the colon, and that these could be traced to specific OTUs on each side.

Next, we built a Random Forest model to differentiate the proximal and distal luminal samples. The model performed best when distinguishing the proximal versus distal mucosa (Fig. 4B, AUC = 0.850) whereas the model to differentiate between the proximal versus distal lumen performed poorly (AUC = 0.580). The OTUs included in the model differentiating the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ (Fig. 6A). The model that attempted to separate the the proximal and distal lumen included OTUs affiliated with the _Bacteroides_, _Clostridium IV_ and _Oscillibacter_ (Fig. 6B). Interestingly, _Anaerococcus_ and _Finegoldia_ were distinct between the mucosa and lumen and also helped to differentiate between the proximal and distal sides.

>>>>

---
title: "Spatial variation of the native colon microbiota in healthy adults"
csl: aacr.csl
fontsize: 12pt
output:
  word_document: default
  pdf_document:
    includes: null
    keep_tex: yes
  html_document: default
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
- \renewcommand{\familydefault}{\sfdefault}
- \usepackage{graphicx}
geometry: margin=1.0in
sansfont: Arial
bibliography: references.bib
---
\vspace{8mm}

@@

Word count: 4,205/5,000

@@

```{r datasets, echo=FALSE}

meta_file <- read.table(file='../data/raw/kws_metadata.tsv', header = T)
shared_file <- read.table(file='../data/mothur/kws_final.an.shared', sep = '\t', header=T, row.names=2)
tax_file <- read.table(file='../data/mothur/kws_final.an.cons.taxonomy', sep = '\t', header=T, row.names=1)
shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
subs_file <- read.table(file='../data/mothur/kws_final.an.0.03.subsample.shared', sep = '\t', header = T, row.names=2)

@@

```{r randomforest, echo=FALSE}
#make OTU abundance file
#Create df with relative abundances
shared_file <- subset(shared_file, select = -c(numOtus, label))
shared_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
rel_abund <- 100*shared_file/unique(apply(shared_file, 1, sum))

#do rel abund calcs for subsampled
subs_file <- subset(subs_file, select = -c(numOtus, label))
subs_meta <- merge(meta_file, shared_file, by.x='group', by.y='row.names')
subs_abund <- 100*subs_file/unique(apply(subs_file, 1, sum))

#Create vector of OTUs with median abundances >1%
OTUs_1 <- apply(rel_abund, 2, max) > 1
OTU_list <- colnames(rel_abund)[OTUs_1]

OTUs_sub <- apply(subs_abund, 2, max) > 1
OTU_list_sub <- colnames(subs_abund)[OTUs_sub]

#get df of just top OTUs
rel_abund_top <- rel_abund[, OTUs_1]
rel_meta <- merge(meta_file, rel_abund_top, by.x='group', by.y="row.names")

subs_abund_top <- subs_abund[, OTUs_sub]
subs_meta <- merge(meta_file, subs_abund_top, by.x='group', by.y="row.names")

seed <- 1
n_trees <- 2001

left_optimized_results <- read.table(file = '../data/process/left_optimized.tsv', sep = '\t')
right_optimized_results <- read.table(file = '../data/process/right_optimized.tsv', sep = '\t')

bowel_optimized_results <- read.table(file = '../data/process/bowel_optimized.tsv', sep = '\t')
stool_optimized_results <- read.table(file = '../data/process/stool_optimized.tsv', sep = '\t')

##############
# make test datasets for each

testsubS <- subset(subs_meta, location %in% c("RS", "LS"))
testsubS$location <- factor(testsubS$location)
levels(testsubS$location) <- c(1:length(levels(testsubS$location))-1)

testsubM <- subset(subs_meta, location %in% c("RB", "LB"))
testsubM$location <- factor(testsubM$location)
levels(testsubM$location) <- c(1:length(levels(testsubM$location))-1)

testsubR <- subset(subs_meta, location %in% c("RB", "RS"))
testsubR$location <- factor(testsubR$location)
levels(testsubR$location) <- c(1:length(levels(testsubR$location))-1)

testsubL <- subset(subs_meta, location %in% c("LB", "LS"))
testsubL$location <- factor(testsubL$location)
levels(testsubL$location) <- c(1:length(levels(testsubL$location))-1)

left_roc <- roc(testsubL$location ~ left_optimized_results$one)
right_roc <- roc(testsubR$location ~ right_optimized_results$one)

bowel_optimized_results <- bowel_optimized_results[-13,]

muc_roc <- roc(testsubM$location ~ bowel_optimized_results$one)
stool_roc <- roc(testsubS$location ~ stool_optimized_results$one)

#print AUCs
auc_muc <- auc(muc_roc)
auc_stool <- auc(stool_roc)

auc_right <- auc(right_roc)
auc_left <- auc(left_roc)

@@

The microbiome has been implicated in the development of colorectal cancer (CRC) and inflammatory bowel diseases (IBD). The specific traits of these diseases vary along the axis of the digestive tract. Further, variation in the structure of the gut microbiota has been associated with both diseases. Here we profiled the microbiota of the healthy proximal and distal mucosa and lumen to better understand how bacterial populations vary along the colon. We used a two-colonoscope approach to sample proximal and distal mucosal and luminal contents from the colons of 20 healthy subjects that had not undergone any bowel preparation procedure. The biopsies and home-collected stool were subjected to 16S rRNA gene sequencing and Random Forest classification models were built using taxa abundance and location to identify microbiota specific to each site. The right mucosa and lumen had the most similar community structures of the five sites we considered from each subject. The distal mucosa had higher relative abundance of _Finegoldia, Murdochiella, Peptoniphilus, Porphyromonas_ and _Anaerococcus_. The proximal mucosa had more of the genera _Enterobacteriaceae, Bacteroides_ and _Pseudomonas_. The classification model performed well when classifying mucosal samples into proximal or distal sides (AUC = `r round(auc_muc, digits=3)`). Separating proximal and distal luminal samples proved more challenging (AUC = `r round(auc_stool, digits=3)`) and specific microbiota that differentiated the two were hard to identify. By sampling the unprepped colon, we identified distinct bacterial populations native to the proximal and distal sides. Further investigation of these bacteria may elucidate if and how these groups contribute to different disease processes on their respective sides of the colon.

Words: 233/250

@@

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that were distinct at each site, we constructed several Random Forest models trained using OTU relative abundances. We built the first model to classify the luminal versus mucosal samples for the proximal and distal sides, independently (Fig. 4A). The models performed well when classifying these samples (proximal AUC = `r round(auc_right, digits=3)`, distal AUC = `r round(auc_left, digits=3)`). The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the proximal lumen and mucosa, OTUs affiliated with the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and _Enterobacteraceae_ were included in the best model (Fig. 5A). The model to differentiate between the distal lumen and mucosa included OTUs affiliated with the _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and _Anaerococcus_ (Fig. 5B). These results indicated that there were fine differences between the different sites of the colon, and that these could be traced to specific OTUs on each side.

Next, we built a Random Forest model to differentiate the proximal and distal luminal samples. The model performed best when distinguishing the proximal versus distal mucosa (Fig. 4B, AUC = `r round(auc_muc, digits=3)`) whereas the model to differentiate between the proximal versus distal lumen performed poorly (AUC = `r round(auc_stool, digits=3)`). The OTUs included in the model differentiating the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ (Fig. 6A). The model that attempted to separate the the proximal and distal lumen included OTUs affiliated with the _Bacteroides_, _Clostridium IV_ and _Oscillibacter_ (Fig. 6B). Interestingly, _Anaerococcus_ and _Finegoldia_ were distinct between the mucosa and lumen and also helped to differentiate between the proximal and distal sides.

### 2018-02-12 submission/manuscript.tex
reviewer edits completed

The microbiome has been implicated in the development of colorectal
cancer (CRC) and inflammatory bowel diseases (IBD). The specific traits
of these diseases vary along the axis of the digestive tract. Further,
variation in the structure of the gut microbiota has been associated
with both diseases. Here we profiled the microbiota of the healthy
proximal and distal mucosa and lumen to better understand how bacterial
populations vary along the colon. We used a two-colonoscope approach to
sample proximal and distal mucosal and luminal contents from the colons
of 20 healthy subjects that had not undergone any bowel preparation
procedure. The biopsies and home-collected stool were subjected to 16S
rRNA gene sequencing and Random Forest classification models were built
using taxa abundance and location to identify microbiota specific to
each site. The right mucosa and lumen had the most similar community
structures of the five sites we considered from each subject. The distal
mucosa had higher relative abundance of \emph{Finegoldia, Murdochiella,
Peptoniphilus, Porphyromonas} and \emph{Anaerococcus}. The proximal
mucosa had more of the genera \emph{Enterobacteriaceae, Bacteroides} and
\emph{Pseudomonas}. The classification model performed well when
classifying mucosal samples into proximal or distal sides (AUC=0.850).
Separating proximal and distal luminal samples proved more challenging
(AUC=0.580) and specific microbiota that differentiated the two were
hard to identify. By sampling the unprepped colon, we identified
distinct bacterial populations native to the proximal and distal sides.
Further investigation of these bacteria may elucidate if and how these
groups contribute to different disease processes on their respective
sides of the colon.

@@

To identify OTUs that were distinct at each site, we constructed several
Random Forest models trained using OTU relative abundances. We built the
first model to classify the luminal versus mucosal samples for the
proximal and distal sides, independently (Fig. 4A). The models performed
well when classifying these samples (proximal AUC = 0.764, distal AUC =
0.908). The OTUs that were most predictive of each site were identified
by their greatest mean decrease in accuracy when removed from the model.
For distinguishing the proximal lumen and mucosa, OTUs affiliated with
the \emph{Bacteriodes}, \emph{Actinomyces}, \emph{Psuedomonas} and
\emph{Enterobacteraceae} were included in the best model (Fig. 5A). The
model to differentiate between the distal lumen and mucosa included OTUs
affiliated with the \emph{Turicibacter}, \emph{Finegoldia},
\emph{Peptoniphilus} and \emph{Anaerococcus} (Fig. 5B). These results
indicated that there were fine differences between the different sites
of the colon, and that these could be traced to specific OTUs on each
side.

Next, we built a Random Forest model to differentiate the proximal and
distal luminal samples. The model performed best when distinguishing the
proximal versus distal mucosa (Fig. 4B, AUC = 0.850) whereas the model
to differentiate between the proximal versus distal lumen performed
poorly (AUC = 0.580). The OTUs included in the model differentiating the
distal and proximal mucosa included members of the \emph{Porphyromonas},
\emph{Murdochiella}, \emph{Finegoldia}, \emph{Anaerococcus} and
\emph{Peptoniphilus} (Fig. 6A). The model that attempted to separate the
the proximal and distal lumen included OTUs affiliated with the
\emph{Bacteroides}, \emph{Clostridium IV} and \emph{Oscillibacter} (Fig.
6B). Interestingly, \emph{Anaerococcus} and \emph{Finegoldia} were
distinct between the mucosa and lumen and also helped to differentiate
between the proximal and distal sides.

>>>>

The microbiome has been implicated in the development of colorectal
cancer (CRC) and inflammatory bowel diseases (IBD). The specific traits
of these diseases vary along the axis of the digestive tract. Further,
variation in the structure of the gut microbiota has been associated
with both diseases. Here we profiled the microbiota of the healthy
proximal and distal mucosa and lumen to better understand how bacterial
populations vary along the colon. We used a two-colonoscope approach to
sample proximal and distal mucosal and luminal contents from the colons
of 20 healthy subjects that had not undergone any bowel preparation
procedure. The biopsies and home-collected stool were subjected to 16S
rRNA gene sequencing and Random Forest classification models were built
using taxa abundance and location to identify microbiota specific to
each site. The right mucosa and lumen had the most similar community
structures of the five sites we considered from each subject. The distal
mucosa had higher relative abundance of \emph{Finegoldia, Murdochiella,
Peptoniphilus, Porphyromonas} and \emph{Anaerococcus}. The proximal
mucosa had more of the genera \emph{Enterobacteriaceae, Bacteroides} and
\emph{Pseudomonas}. The classification model performed well when
classifying mucosal samples into proximal or distal sides (AUC = 0.808).
Separating proximal and distal luminal samples proved more challenging
(AUC = 0.599) and specific microbiota that differentiated the two were
hard to identify. By sampling the unprepped colon, we identified
distinct bacterial populations native to the proximal and distal sides.
Further investigation of these bacteria may elucidate if and how these
groups contribute to different disease processes on their respective
sides of the colon.

@@

To identify OTUs that were distinct at each site, we constructed several
Random Forest models trained using OTU relative abundances. We built the
first model to classify the luminal versus mucosal samples for the
proximal and distal sides, independently (Fig. 4A). The models performed
well when classifying these samples (proximal AUC = 0.716, distal AUC =
0.862). The OTUs that were most predictive of each site were identified
by their greatest mean decrease in accuracy when removed from the model.
For distinguishing the proximal lumen and mucosa, OTUs affiliated with
the \emph{Bacteriodes}, \emph{Actinomyces}, \emph{Psuedomonas} and
\emph{Enterobacteraceae} were included in the best model (Fig. 5A). The
model to differentiate between the distal lumen and mucosa included OTUs
affiliated with the \emph{Turicibacter}, \emph{Finegoldia},
\emph{Peptoniphilus} and \emph{Anaerococcus} (Fig. 5B). These results
indicated that there were fine differences between the different sites
of the colon, and that these could be traced to specific OTUs on each
side.

Next, we built a Random Forest model to differentiate the proximal and
distal luminal samples. The model performed best when distinguishing the
proximal versus distal mucosa (Fig. 4B, AUC = 0.808) whereas the model
to differentiate between the proximal versus distal lumen performed
poorly (AUC = 0.599). The OTUs included in the model differentiating the
distal and proximal mucosa included members of the \emph{Porphyromonas},
\emph{Murdochiella}, \emph{Finegoldia}, \emph{Anaerococcus} and
\emph{Peptoniphilus} (Fig. 6A). The model that attempted to separate the
the proximal and distal lumen included OTUs affiliated with the
\emph{Bacteroides}, \emph{Clostridium IV} and \emph{Oscillibacter} (Fig.
6B). Interestingly, \emph{Anaerococcus} and \emph{Finegoldia} were
distinct between the mucosa and lumen and also helped to differentiate
between the proximal and distal sides.
